D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-21 ootavacuous.tex
Alan Stern February 21 feedback, Section 8.2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
memory model.
The ``order-preserving'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering only of quasi volatile accesses to the same object,
but such reordering is in any case directly forbidden by the standard.
Referring also to the ``onto'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses},
implementations must not behave as if the source code has
additional:

@@

\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Do Not Duplicate Atomic Loads}
\label{lst:Do Not Duplicate Atomic Loads}
\end{listing}

However, even duplicating an existing atomic load can lead to incorrect
results.
To see this, please refer to
Listing{lst:Do Not Duplicate Atomic Loads}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} might
not be prepared to deal with.

Therefore, C++ implementations should not invent loads, and should
not even duplicate them.

@@

>>>>

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store may
execute only after the committing of all loads whose return values are
used to compute that store's address or value.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
memory model.
The ``order-preserving'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering only of quasi volatile accesses to the same object,
but such reordering is in any case directly forbidden by the standard.
Referring also to the ``onto'' bullet from
Section~\ref{sec:Volatile and Quasi Volatile Accesses},
implementations should not behave as if the source code has
additional:

@@

\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Do Not Duplicate Atomic Loads}
\label{lst:Do Not Duplicate Atomic Loads}
\end{listing}

However, even duplicating an existing atomic load can lead to incorrect
results.
To see this, please refer to
Listing~\ref{lst:Do Not Duplicate Atomic Loads}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} might
not be prepared to deal with.

Therefore, C++ implementations should avoid inventing atomic loads,
and in fact should avoid even duplicating them.

@@

This provides an example of the ``one-to-one'' bullet of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
